---
title: "Modelling Assignment"
author: "Tom Delany"
date: "2026-03-08"
output: html_document
---


```{r}
library(dplyr)
library(gt)
library(mice)
library(tree)
library(ggplot2)
```



```{r}
framing <- read.csv("framingham_heart_study.csv")

dim(framing)
length(is.na(framing))

head(framing)

missing_table <- data.frame(
  Variable = names(framing),
  Missing_Count = colSums(is.na(framing)),
  Missing_Proportion = colMeans(is.na(framing)),
  Missing_Percent = colMeans(is.na(framing)) * 100
)

missing_table

sum(framing$TenYearCHD == 1)
```



This analysis will evaluate the predictive performance and potential clinical usefulness of different modelling approaches for estimating the 10-year risk of coronary heart disease (CHD) using the Framingham dataset.

The following sections will roughly follow the slides in my final presentation.

Goals of the project:

Handle missing values (First through ascertain of the type of missingness (MAR,MCAR,NMAR)). Missing values in predictor variables will first be addressed using imputation techniques to minimise bias and retain all observations. Following imputation, both traditional statistical learning methods (such as logistic regression) and machine learning approaches (including decision trees and random forests) will be trained to predict the binary outcome TenYearCHD.

Model performance will be compared using discrimination metrics such as the area under the receiver operating characteristic curve (AUC) and calibration assessments. Finally, decision curve analysis will be used to evaluate the net clinical benefit of each model across a range of risk thresholds, allowing assessment of which modelling approach may offer the greatest potential utility in a clinical decision-making context.


Missing data - why must we handle it:
Missing data must be addressed in this analysis because several predictors in the dataset contain incomplete observations. If these missing values were ignored and a complete case analysis were used, all subjects with any missing predictor would be excluded, which discards a substantial amount of available information and reduces statistical efficiency.
For example, even when only a small proportion of predictor values are missing, complete case analysis may exclude a large proportion of subjects from the model, leading to less reliable estimates and reduced power.
In addition, missing data may introduce selection bias if individuals with missing values differ systematically from those with complete data. Imputation methods therefore allow the use of information from partially observed records while reducing bias and improving the reliability of prediction models. 

## Missingness Diagnostics

Investigating Missingness:
Missing data can arise through different mechanisms, including Missing Completely at Random (MCAR), where missingness occurs purely by chance; Missing at Random (MAR), where missingness is related to other observed variables in the dataset; and Missing Not at Random (MNAR), where the probability of missingness depends on the unobserved value itself or other unmeasured factors.

```{r}
missing_table %>%
  arrange(desc(Missing_Count))
```

```{r}
png("missing_pattern.png", width = 1400, height = 900, res = 150)
md.pattern(framing, plot = TRUE)
dev.off()
```

```{r}
# Variables with missingness in this dataset
partial_vars <- names(which(colSums(is.na(framing)) > 0))
partial_vars
```

## Single and Multiple Imputation Using `mice` (Ref: Van Buuren 2011)

First we split:
```{r}
set.seed(214)

train_index <- sample(1:nrow(framing), 0.7 * nrow(framing))

train_data <- framing[train_index, ]
test_data  <- framing[-train_index, ]
```


Now we will do multiple imputation using the `mice()` function in the `mice` library. NB `mice` stands for multivariate imputation by chained equations. We include `TenYearCHD` as a predictor in imputation models but do not impute `TenYearCHD`, so we can analyse each imputed dataset afterwards.

```{r}
train_data <- train_data %>%
  mutate(
    BPMeds = factor(BPMeds),                       # binary with missingness
    education = ordered(education, levels = 1:4)   # ordinal with missingness
  )

test_data <- test_data %>%
  mutate(
    BPMeds = factor(BPMeds),                       # binary with missingness
    education = ordered(education, levels = 1:4)   # ordinal with missingness
  )

```

```{r}
framing_to_impute <- train_data

init <- mice(framing_to_impute, maxit = 0, print = FALSE)
meth <- init$method
pred <- init$predictorMatrix

# Explicit methods
meth[] <- ""
meth["cigsPerDay"] <- "norm"
meth["BPMeds"] <- "logreg"
meth["education"] <- "polr"
meth["totChol"] <- "norm"
meth["BMI"] <- "norm"
meth["heartRate"] <- "norm"
meth["glucose"] <- "norm"
meth["TenYearCHD"] <- ""  # do not impute outcome

# Ensure outcome is not imputed but can be used to predict other variables
pred["TenYearCHD", ] <- 0

meth
pred
```

Train Imputation
```{r}
framing.imputed.train <- mice(
  data = train_data,
  m = 1,
  method = meth,
  predictorMatrix = pred,
  seed = 214,
  maxit = 5,
  print = FALSE
)

train_complete <- complete(framing.imputed.train)
```

Test Imputation

```{r}
#drop tenyearCHd being used in test impuation
pred_test <- pred
pred_test[, "TenYearCHD"] <- 0

framing.imputed.test <- mice(
  data = test_data,
  m = 1,
  method = meth,
  predictorMatrix = pred_test,
  seed = 214,
  maxit = 5,
  print = FALSE
)

test_complete <- complete(framing.imputed.test)
```


```{r}
# Convergence diagnostics
plot(framing.imputed.train)
```
```{r}
# Convergence diagnostics
plot(framing.imputed.test)
```

```{r}


train_complete %>%
  ggplot() +
  geom_density(aes(x = glucose), col = "red", alpha = 0.01) +
  geom_density(data = framing, aes(x = glucose), col ="blue")
  
```

```{r}
#Changing to factor so tree isnt a regression but classification tree
train_complete$TenYearCHD <- factor(train_complete$TenYearCHD)
test_complete$TenYearCHD  <- factor(test_complete$TenYearCHD)
```



Naive tree
```{r}
library(tree)

tree.model <- tree(TenYearCHD ~ ., data = train_complete)

tree.pred <- predict(tree.model, test_complete, type = "class")

tree_cm <- table(Predicted = tree.pred, Actual = test_complete$TenYearCHD)
tree_cm

tree_accuracy <- mean(tree.pred == test_complete$TenYearCHD)
tree_misclass <- 1 - tree_accuracy
tree_sensitivity <- tree_cm["1", "1"] / sum(tree_cm[, "1"])
tree_specificity <- tree_cm["0", "0"] / sum(tree_cm[, "0"])

c(
  Accuracy = tree_accuracy,
  Misclassification = tree_misclass,
  Sensitivity = tree_sensitivity,
  Specificity = tree_specificity
)
```

```{r}
plot(tree.model)
text(tree.model, pretty = 0, cex = 0.7)
```


Random Forest
```{r}

library(randomForest)
Framing.rf <- randomForest(TenYearCHD ~ ., 
                        data = train_complete, 
                        importance = TRUE,
                        proximity = TRUE)
```

```{r}
rf_pred <- predict(Framing.rf, newdata = test_complete)
rf_cm <- table(Predicted = rf_pred, Actual = test_complete$TenYearCHD)
rf_cm

rf_accuracy <- mean(rf_pred == test_complete$TenYearCHD)
rf_misclass <- 1 - rf_accuracy
rf_sensitivity <- rf_cm["1", "1"] / sum(rf_cm[, "1"])
rf_specificity <- rf_cm["0", "0"] / sum(rf_cm[, "0"])
rf_oob_error <- Framing.rf$err.rate[nrow(Framing.rf$err.rate), "OOB"]

c(
  Accuracy = rf_accuracy,
  Misclassification = rf_misclass,
  Sensitivity = rf_sensitivity,
  Specificity = rf_specificity,
  OOB_Error = rf_oob_error
)
```

```{r}
plot(Framing.rf)
varImpPlot(Framing.rf)
```


Bayesian Additive Regression Trees (BART)
```{r}
library(BART)

x_train <- train_complete %>% dplyr::select(-TenYearCHD)
x_test  <- test_complete %>% dplyr::select(-TenYearCHD)
y_train <- as.numeric(as.character(train_complete$TenYearCHD))
y_test  <- as.numeric(as.character(test_complete$TenYearCHD))

set.seed(1)
bartfit <- pbart(x.train = x_train, y.train = y_train, x.test = x_test)
```

```{r}
# Predicted probability of CHD=1
bart_prob <- if (!is.null(bartfit$prob.test.mean)) bartfit$prob.test.mean else bartfit$yhat.test.mean
bart_pred <- factor(ifelse(bart_prob >= 0.5, 1, 0), levels = c(0, 1))

bart_cm <- table(Predicted = bart_pred, Actual = test_complete$TenYearCHD)
bart_cm

bart_accuracy <- mean(bart_pred == test_complete$TenYearCHD)
bart_misclass <- 1 - bart_accuracy
bart_sensitivity <- bart_cm["1", "1"] / sum(bart_cm[, "1"])
bart_specificity <- bart_cm["0", "0"] / sum(bart_cm[, "0"])

c(
  Accuracy = bart_accuracy,
  Misclassification = bart_misclass,
  Sensitivity = bart_sensitivity,
  Specificity = bart_specificity
)
```

```{r}
# Variable usage across posterior trees
ord <- order(bartfit$varcount.mean, decreasing = TRUE)
bartfit$varcount.mean[ord]
```


Ridge regression (benchmark)
```{r}
library(glmnet)

x_train_mm <- model.matrix(TenYearCHD ~ ., data = train_complete)[, -1]
x_test_mm  <- model.matrix(TenYearCHD ~ ., data = test_complete)[, -1]
y_train_num <- as.numeric(as.character(train_complete$TenYearCHD))

set.seed(214)
ridge_cv <- cv.glmnet(
  x = x_train_mm,
  y = y_train_num,
  family = "binomial",
  alpha = 0
)

ridge_prob <- as.numeric(predict(ridge_cv, newx = x_test_mm, s = "lambda.min", type = "response"))
ridge_pred <- factor(ifelse(ridge_prob >= 0.5, 1, 0), levels = c(0, 1))

ridge_cm <- table(Predicted = ridge_pred, Actual = test_complete$TenYearCHD)
ridge_cm

ridge_accuracy <- mean(ridge_pred == test_complete$TenYearCHD)
ridge_misclass <- 1 - ridge_accuracy
ridge_sensitivity <- ridge_cm["1", "1"] / sum(ridge_cm[, "1"])
ridge_specificity <- ridge_cm["0", "0"] / sum(ridge_cm[, "0"])

c(
  Accuracy = ridge_accuracy,
  Misclassification = ridge_misclass,
  Sensitivity = ridge_sensitivity,
  Specificity = ridge_specificity,
  Lambda = ridge_cv$lambda.min
)
```






